Hunting experience shapes individual foraging specialization and predator-prey interactions in an online videogame

### 2023-03-29 manuscript/Chapter2-manuscript.Rmd
integrate most of Francesca's comments

Individual variation in predator foraging behaviour is increasingly recognized as a major driver of trophic interactions and community dynamics [@Griffen.etal2012;@Michalko.Pekar2016;@Moran.etal2017;@Michalko.etal2021]. Indeed, predator populations often comprise assemblages of individuals specializing in different foraging tactics or on different resources irrespective of sexual, morphological, or age-related differences [@Estes.etal2003;@Tinker.etal2008;@kernaleguenDeterminantsIndividualForaging2015;@Phillips.etal2017]. A growing body of evidence suggests that ecological interactions, such as predator-prey interactions, can drive such individual foraging specialization [@Araujo.etal2011;@toscanoPersonalityForagingBehavior2016]. When they hunt, predators often use techniques that are fine-tuned to the type of prey that they encounter [@Davoren.etal2003;@Estes.etal2003;@Woo.etal2008;@Courbin.etal2018], and their capacity to use them effectively is contingent on periods of extensive practice (i.e. experience). While hunting experience may be essential to maintain or improve foraging success, we have few empirical assessments of its role in predator foraging specialization [@Dukas2019] and its consequences on predator-prey interactions.

The development of foraging tactics is crucial for young predators to reach adulthood and survive [@Phillips.etal2017;@Heithaus.etal2018]. Theory predicts that foraging specialization may emerge via learning, limitations in memorizing multiple complex hunting skills, and expertise [@Tinker.etal2009;@Dukas2019]. @Dukas2019 defines expertise as the characteristics, skills, and knowledge allowing individuals with extensive experience to outperform novices on complex tasks. This body of work suggests that the development of expertise through extensive practice is an optimizing process that should promote foraging specialization. Empirical studies on human and non-human hunters show that experience optimizes the efficiency (e.g. search and handling times, return rates) of their foraging tactics potentially via associative images or reliance on prey and environmental cues [@Edwards.Jackson1994;@Morse2000;@macdonaldCrossculturalComparisonLearning2007;@Reid.etal2010;@Wilson-Rankin2015]. Such optimization may thus reinforce the use of the same tactic (i.e. specialization) if its success is constant each time a prey is encountered. It may also be costly to attempt different hunting tactics by trial and error when prey are scarce or highly unpredictable [@Dukas1998;@Estes.etal2003;@Mery.Burns2010]. An alternative mechanism is that the costs/risks faced by hunters when switching foraging tactics may be offset by gaining experience and information on their prey, leading to increased individual foraging flexibility [@Stephens1993;@Ishii.Shimada2010;@Mery.Burns2010;@Kelley.Magurran2011;@Snell-Rood2013]. To develop proper responses to fluctuating resources, individuals would need to sample their environment broadly during an extensive period of time, followed by higher performance at later stages of development [reviewed in @Snell-Rood2013]. However, predators can differ in the challenges that they face during their lifetime, which could lead to both mechanisms operating at the same time within a predator population. For instance, some individuals may constantly encounter faster prey, which are proven to be more difficult to hunt [@Walker.etal2005;@Kelley.Magurran2011;@FraserFranco.etal2022].

There is currently a lack of consensus on the fitness advantages of specialized vs flexible foraging. Importantly, we have limited information on the ecological contexts - except for competition - that favour specialization over flexibility. This is reflected in the literature showing contrasting results in the links between specialization and fitness. For instance, some studies report increasing benefits of specialization [@Pintor.etal2014;@Patrick.Weimerskirch2014a;@vandenBosch.etal2019], some report that flexible foraging has greater benefits [@Paull.etal2012;@Manlick.etal2021], and others find equal benefits depending on timescales [@Woo.etal2008;@Potier.etal2015]. While quantifying the fitness consequences of specialization is a daunting task, in predator-prey systems, some clues indicate that fluctuations in the predictability of prey encounters throughout a predator's lifetime may be a key factor [@Weimerskirch2007;@Woo.etal2008;@Chang.etal2017;@Phillips.etal2017;@Courbin.etal2018]. The resource-predictability hypothesis argues that when resources are predictable, individual specialists should have higher delivery rates by reducing the energy and time required to search for and handle prey. In contrast, individual generalists should benefit when resources fluctuate, as fine adjustments to resources are key for a predator's success and survival [@Karkarey.etal2017;@Holm.etal2019;@Santoro.etal2019]. At the phenotypic interface, empirical evidence shows that predators often match, among other types of traits [see @Kishida.etal2006;@Hanifin.etal2008;@Brousseau.etal2018;@Reimche.etal2020], their locomotor and behavioural traits to those of their prey [@McGhee.etal2013;@Bro-Jorgensen2013;@Chang.etal2017;@Szopa-Comley.Ioannou2022]. Yet, a recurring question that emerges from these studies is if and how the behavioural adjustments of predators reflect individual differences in experience/learning [@Kelley.Magurran2011]. Thus, uncovering the role of learning in predator-prey systems where interactions are directly monitored would enable researchers to better predict the behavioural decisions and the success of predators when they are hunting.

The integration of individual behavioural variation in the study of predator-prey interactions has gained traction in recent years, with empirical studies revealing important consequences for habitat use, functional responses, prey choice, and foraging rate [@Kobler.etal2009;@toscanoTraitmediatedFunctionalResponses2014;@Patrick.Weimerskirch2014;@Matsumura.Miyatake2022a]. However, an important and recurring challenge impeding research on predator-prey behavioural interactions, at the individual level, is the need to collect data simultaneously on both the predator and prey. We recently demonstrated with behavioural data from an online predator-prey videogame that virtual systems can overcome this challenge and help uncover the mechanisms that shape predator-prey interactions [@FraserFranco.etal2022]. For instance, we found that the classical locomotor crossover hypothesis [@Huey.Pianka1981] applied only to cursorial hunters [see also: @Matsumura.Miyatake2022a], suggesting that virtual systems could be useful in identifying general ecological patterns, which might in turn help to advance the current gaps in predator-prey research. Other studies on virtual predator-prey systems show that predation regimes can drive individual variation in risk perception [@Beauchamp2020], that familiarity between prey has a positive indirect effect on survival [@Cere.etal2021], and that prey face contrasting natural and social selection regimes (Santostefano et al. in prep). Because virtual predator-prey systems generate large volumes of data on interacting players throughout their lifetime in the game, they offer the opportunity to tackle fundamental questions about the role of experience and prey behaviour on individual predator foraging specialization along with their potential fitness consequences.

In this study, we test how hunting experience shapes predator foraging specialization using individual behavioural data from players in the online videogame *Dead by Daylight* (*DBD*). *DBD* simulates a direct predator-prey interaction, where one predator player hunts four prey players in different virtual environments. The data grants a high degree of precision on the behavioural interaction, as the behaviour of both the predator and the four prey is monitored simultaneously in each trial. First, we investigate how predators develop their individual hunting expertise. We hypothesize that predators should differ in the development of their expertise, partly because they encounter varying levels of difficulty with the prey that they pursue. For example, a predator may face greater difficulty than other individuals if it more often encountered prey that were elusive. Second, we test the hypothesis that experience will shape foraging specialization. If experience reduces the costs of switching between hunting tactics, we predict that the predator population should become more flexible. Alternatively, if experience enables the refinement of the hunting tactics, then the predator population may instead specialize. Otherwise, if both mechanisms operate at the same time, then we should see no change with experience at the population level, but differences in specialization among individuals should increase. Third, we evaluate how predator foraging specialization emerges from behavioural interactions with prey. Whether predators specialize or not with experience should depend on the behaviour of their prey. We expect that predators that experienced more predictable encounters with their prey will specialize, while predators that experienced unpredictable encounters with their prey should adopt a flexible hunting strategy. If we detect such prey-dependent fine-tuning, then specialist and flexible hunters should attain equal success.

@@

*DBD* is a survival asymmetric (i.e. a game where the gameplay  mechanics differ between two groups) multiplayer online game developed by Behaviour Interactive Inc, in which players can play either as a predator or a prey. The objective of the predator is to hunt and capture the four prey across a virtual environment. The objective of the four prey is to search for resources while avoiding the predator. The resources are in the form of power generators that, once they are all activated, will enable the prey to escape through one of two exit doors. The composition of the predator and prey group for a match is determined by a skill-based matchmaking algorithm. A match ends when the predator kills all the prey available (i.e. that have not escaped), or when the last remaining prey escapes the virtual environment. Each player, predator or prey, can choose an avatar with abilities that encourage specific play styles (e.g. bold vs cautious prey, or ambush vs roaming predator). During our study period, the game offered 23 predator avatars. The virtual environment where matches take place is composed of fixed and procedurally generated habitat components, such as vegetation, mazes, and buildings. Some of these environments are larger than others, with varying structural complexity. However, we have previously shown that predators display only minimal changes in behaviour and hunting success across the environments, probably due to a game feature enabling predators to have visual cues of the generators [@FraserFranco.etal2022]. There were `r toString(format(length(unique(dat$environment_id))))` virtual game environments available for play during our study period.

In *DBD*, some individual predators specialize on hunting at high speeds and covering space in the environment, while others prefer to stalk and ambush their prey [@FraserFranco.etal2022]. Individuals are also flexible in the use of these tactics by switching between them across matches. The expression of these tactics and their success is also shaped by the behaviour of the prey group. The prey need to forage for resources while paying attention to the predator to avoid being detected and chased. Some prey contribute to the group's success by healing or helping others escape the predator, while others play alone and attempt to escape by themselves [@Cere.etal2021]. Predators must learn how the prey behave and then decide how best to capture them. Thus, *DBD* simulates a highly dynamic system where both predators and prey must constantly adjust to each other to be successful.

@@

The videogame company provided data that spanned a period of 6 months of gameplay recorded for every player. The first recorded match was played on `r toString(format(min(dat$date)))` and the last one on `r toString(format(max(dat$date)))`. We cleaned and filtered the raw data to produce a dataset appropriate for our analyses. We analyzed only matches where players did not know each other (i.e. "Online" mode). We filtered any matches where players were inactive, such as when speed values were equal to, or very close to, zero. Moreover, we used our knowledge of the game to remove any matches where players were potentially hacking, or not playing the game has it is supposed to be played. We then partitioned the player population by total experience, and sampled players that played 300 matches or more. To optimize the representation of gameplay experience at advanced levels, we set the maximum number of matches at 500 for these players. For instance, there could be a large difference in gameplay between 300-500 and 500-1000 matches. Players that played more than 500 matches represented ~2% of the population.

Our population consists of `r toString(format(length(unique(dat$predator_id))))` predator players with a total record of `r toString(formatC(length(unique(dat$match_id)), big.mark = " "))` matches. The predator-players' experience varied between `r toString(format(min(dat$total_xp_pred)))` and `r toString(format(max(dat$total_xp_pred)))` matches played. These matches lasted between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)`-min (mean = `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)`-min). The following information is collected and reported for every match : the player's anonymous ID, its avatar, the game environment, the predator-player's experience along with its speed, and the average speed of the group of prey it encountered.

We recognize that we could have introduced a bias in our analyses since we sampled individuals who stayed in the game for at least 300 matches. For instance, these players may already be used to play videogames, and thus, could differ in the way they play at early stages of their experience in the game. To ensure that this was not the case, we randomly sampled players that played between 20 and 50 matches, between 51 and 100 matches, and between 101 and 300 matches. We then took the first 20 matches played by these players, including those from our sampled population, and compared their average behaviour using a Bayesian hierarchical linear model. We found that neither of the four groups differed in their average speed (Table SX), which gives us confidence that our player population was not biased.

@@

We analyzed the predator's average speed and the average speed of the prey encountered by the predator. The predator's average speed is measured as the average distance traveled per second during a match ($\overline{x}$ = `r toString(format(round(mean(dat$pred_speed), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$pred_speed), digits = 2), nsmall = 2))` m/s). We measured the preys' average speed as the average travel speed of the four individual prey within a match ($\overline{x}$ = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s). We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). Lastly, we defined the predator's cumulative experience as the number of matches played prior to the match being monitored. For example, the first match of a player would have a cumulative experience value of 0, while the tenth match would have a value of 9.

@@

We used the default number of basis functions (K) in "brms" for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id$) followed a Gaussian distribution with estimated standard deviation ($id \sim N(0,\ I\ \sigma_{id})$). We used Gaussian priors for the game duration ($N(0,\ 1)$), the intercept ($N(0,\ 0.5)$), and the precision paremeter ($N(2,\ 0.5)$). We employed the default Student t priors for the smoothing parameter ($t(3,\ 0,\ 2.5)$). In all models, we specified four MCMC chains to estimate the parameters, with 1000 posterior samples for each parameter. To do so, we ran 2500 iterations for the first model with a thinning set to eight, and 1500 iterations with a thinning set to four for the second and third models; burn-in was set to 500 iterations in each model. We assessed the convergence of the MCMC chains using trace plots, R-hat diagnostics with a threshold of <1.05, and effective sample sizes (ESS) with a threshold of >100 for the bulk-ESS and tail-ESS [@Vehtari.etal2021a]. After fitting the three models, we proceeded to select the one with the best predictive accuracy using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@Vehtari.etal2017;@Piironen.Vehtari2017;@Vehtari.etal2022].

### Foraging behaviour and success, prey behaviour, and their relationships, at each level of experience

We tested whether foraging behaviour changed with experience, and compared the success of generalists vs specialists, using a multivariate, double-hierarchical generalized linear model (MDHGLM) [@leeDoubleHierarchicalGeneralized2006;@cleasbyQuantifyingPredictabilityBehaviour2015;@westneatBiologyHiddenResidual2014;@Mitchell.etal2016a;@ODea.etal2022]. We categorized the predators by their experience based on the number of matches they played (i.e. their cumulative experience). We assigned the status of "novice" to any predator-player that had a cumulative experience below 100 matches. For example, a predator player that played 500 matches in total would be labelled as a novice in its first 100 matches. Following this logic, we labelled predators as "intermediate" hunters when they had a cumulative experience between 100 and 299 and matches. For the third and last group, we defined predators as "advanced" hunters when they had a cumulative experience equal or above 300, with a maximum of 499 cumulated matches.

For each level of experience (i.e. novice, intermediate, advanced), the model estimated the mean speed of every individual (which we call the mean part of the model) and, by using a heterogeneous structure of the residuals, the intra-individual standard deviation of speed for every individual (which we call the dispersion part of the model). Thus, in the mean part of the model, we could estimate among individual differences in the foraging tactic used by predators (i.e. fast or slow), while in the dispersion part of the model, we could estimate among individual differences in foraging specialization/flexibility (i.e. intra-individual variability). We followed the same structure for the preys' speed to estimate among individual differences in the mean and standard deviation of speed of the prey encountered by predators. For hunting success, we only modeled the mean part of the equation to estimate among individual differences in mean hunting success. For the predator and the prey's speed, we controlled for the average rank of the prey that they encountered on both the mean and dispersion part of the equation. The ranking system in *DBD* was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank). We used the mean rank of the four prey and applied a square-root transformation. For the predator's hunting success, we controlled for the square-root of the duration of the match.

We estimated the correlation between the individual random effect on the mean and on the dispersion within each trait (e.g. $cor(\sigma_{id_{m}\ pred}\ \sigma_{id_{d}\ pred})$) and among each trait (e.g. $cor(\sigma_{id_{m}\ pred}\ \sigma_{id_{d}\ prey})$) within a level of experience. In the first case, this allows us to assess, for example, whether predators that are on average slower/faster are more specialized or flexible. In the second case, the correlations enables us to estimate, for example, whether predators that encountered prey that were faster/slower adopted a slower/faster (or more specialized/flexible) tactic, or if slower/faster predators had lower/greater hunting success. As a complement, the model also estimated correlations among the different levels of experience (e.g. $cor(\sigma_{id_{m}\ pred\ novice}\ \sigma_{id_{d}\ pred\ intermediate})$). The latter "character-state" approach [@Via.Lande1985;@viaAdaptivePhenotypicPlasticity1995], enables us to investigate whether predators express similar behaviours across each of three developmental stages. We present the character-state analyses as additional material in the Appendix S2 because they were not part of our initial objectives.

@@

We parameterized equations (1) and (2) with a Gaussian distribution family (predator and prey speed) and used a beta-binomial distribution family for equation (3) (predator hunting success). We used Gaussian priors for the prey rank ($N(2.5,\ 1)$) and game duration ($N(10,\ 2)$) on the mean part of the model, and the default priors in "brms" for the dispersion part. We used the default Student t priors in "brms" for the intercepts on the mean and dispersion part of the equation. For hunting success, we applied a Gaussian prior on the precision parameter ($N(2,\ 1)$). For every predicted variable, we applied a half-Gaussian on the random effects for the mean part of the model ($N(0,\ 1)$), and the default "brms" Student t prior for the dispersion part of the model. We parametrized the model to run four MCMC chains with 1000 posterior samples for each parameter by completing 2500 iterations having a thinning set to 8, with the first 500 iterations used as warmups. We assessed the convergence of the MCMC chains using trace plots, R-hat diagnostics with a threshold of <1.05, and effective sample sizes (ESS) with a threshold of >100 for the bulk-ESS and tail-ESS [@Vehtari.etal2021a]. We also performed posterior predictive checks which showed an adequate fit of the model (for details, see https://github.com/quantitative-ecologist/experience-hunting-tactics). We additionally report variance partitioning and coefficients of variations in the supporting information (Appendix S1).

@@

# AKNOWLEDGEMENTS

>>>>

Individual variation in predator foraging behaviour is increasingly recognized as a major driver of trophic interactions and community dynamics [@Griffen.etal2012;@Michalko.Pekar2016;@Moran.etal2017;@Michalko.etal2021]. Indeed, predator populations often comprise assemblages of individuals specializing in different foraging tactics or on different resources irrespective of sexual, morphological, or age-related differences [@Estes.etal2003;@Tinker.etal2008;@kernaleguenDeterminantsIndividualForaging2015;@Phillips.etal2017]. A growing body of evidence suggests that ecological interactions such as predator-prey interactions can drive this individual foraging specialization [@Araujo.etal2011;@toscanoPersonalityForagingBehavior2016]. When they hunt, predators often use techniques that are fine-tuned to the type of prey that they encounter [@Davoren.etal2003;@Estes.etal2003;@Woo.etal2008;@Courbin.etal2018], and their capacity to use them effectively is contingent on periods of extensive practice (i.e. experience). While hunting experience may be essential to maintain or improve foraging success, we have few empirical assessments of its role in predator foraging specialization [@Dukas2019] and its consequences on predator-prey interactions.

The development of foraging tactics is crucial for young predators to reach adulthood and survive [@Phillips.etal2017;@Heithaus.etal2018]. Theory predicts that foraging specialization may emerge via learning, limitations in memorizing multiple complex hunting skills, and expertise [@Tinker.etal2009;@Dukas2019]. @Dukas2019 defines expertise as the characteristics, skills, and knowledge allowing individuals with extensive experience to outperform novices on complex tasks. It is described as the relationship between success and cumulative experience, where extensive practice optimizes success when individuals reach an advanced level of experience. Empirical studies on human and non-human hunters show that experience optimizes the efficiency (e.g. search and handling times, return rates) of their foraging tactics potentially via associative images or reliance on prey and environmental cues [@Edwards.Jackson1994;@Morse2000;@macdonaldCrossculturalComparisonLearning2007;@Reid.etal2010;@Wilson-Rankin2015]. This body of work suggests that such optimization may thus promote individual foraging specialization (e.g. repeatedly using fast attacks) if the success of a hunting tactic is constant each time a prey is encountered. It may also be costly to attempt different hunting tactics by trial and error when prey are scarce or highly unpredictable [@Dukas1998;@Estes.etal2003;@Mery.Burns2010]. An alternative mechanism is that the costs/risks faced by hunters when switching foraging tactics may be offset through extensive practive, as individuals gain experience and information on their prey, leading to increased individual foraging flexibility [@Stephens1993;@Ishii.Shimada2010;@Mery.Burns2010;@Kelley.Magurran2011;@Snell-Rood2013]. To develop proper responses to fluctuating resources, individuals would need to sample their environment broadly during an extensive period of time, leading to greater performance at later stages of development [reviewed in @Snell-Rood2013]. However, predators can differ in the challenges that they face during their lifetime, which could result in both mechanisms operating at the same time within a predator population. For instance, some individuals may constantly encounter faster prey, which are proven to be more difficult to hunt in simulation and empirical studies [@Walker.etal2005;@Kelley.Magurran2011;@Martin.etal2022]. Thus, whether predators specialize or increase their flexibility with experience, learning to adjust their tactic to their prey should be essential to optimize their success.

There is currently a lack of consensus on the fitness advantages of specialized vs flexible foraging [@Phillips.etal2017]. This is reflected in the literature showing contrasting results in the links between specialization and fitness. For instance, some studies report increasing benefits of specialization [@Pintor.etal2014;@Patrick.Weimerskirch2014a;@vandenBosch.etal2019], some report that flexible foraging has greater benefits [@Paull.etal2012;@Manlick.etal2021], and others find equal benefits depending on timescales [@Woo.etal2008;@Potier.etal2015]. This may in part be due to the limited information we have on the ecological contexts favouring specialization over flexibility in nature. However, in predator-prey systems, many studies have outlined that fluctuations in the predictability of prey encounters throughout a predator's lifetime may be a key factor shaping foraging specialization [@Weimerskirch2007;@Woo.etal2008;@Chang.etal2017;@Phillips.etal2017;@Courbin.etal2018]. The resource-predictability hypothesis argues that when resources are predictable (or stable), individual specialists should have higher capture rates by reducing the energy and time required to search for and handle prey. In contrast, individual generalists should benefit when resources fluctuate, as fine adjustments to resources are key for a predator's success and survival [@Karkarey.etal2017;@Holm.etal2019;@Santoro.etal2019]. When predators directly interact with their prey, empirical evidence shows that they often match, among other types of traits [see @Kishida.etal2006;@Hanifin.etal2008;@Brousseau.etal2018;@Reimche.etal2020], their locomotor and behavioural traits to those of their prey [@McGhee.etal2013;@Bro-Jorgensen2013;@Chang.etal2017;@Szopa-Comley.Ioannou2022]. Yet, a recurring question that emerges from these studies is whether and how the behavioural adjustments of predators reflect individual differences in experience/learning [@Kelley.Magurran2011]. Thus, uncovering the role of learning in predator-prey systems where interactions are directly monitored would enable researchers to better predict the behavioural decisions and the success of predators when they are hunting.

The integration of individual behavioural variation in the study of predator-prey interactions has gained traction in recent years, with empirical studies revealing important consequences for habitat use, functional responses, prey choice, and foraging rate [@Kobler.etal2009;@toscanoTraitmediatedFunctionalResponses2014;@Patrick.Weimerskirch2014;@Matsumura.Miyatake2022a]. However, an important and recurring challenge impeding research on predator-prey behavioural interactions, at the individual level, is the need to collect data simultaneously on both the predator and prey. We recently demonstrated with behavioural data from an online predator-prey videogame, called *Dead by Daylight* (*DBD*), that virtual systems can overcome this challenge and help uncover the mechanisms that shape predator-prey interactions [@FraserFranco.etal2022]. For instance, we found that some individual predators hunt at high speeds and cover space in the environment, while others prefer to stalk and ambush their prey. Individuals are also flexible in the use of these tactics by switching between them across matches. The expression of these tactics and their success is also shaped by the speed of the prey, where faster prey are more difficult to hunt. The prey need to forage for resources while paying attention to the predator to avoid being detected and chased. Some prey contribute to the group's success by healing or helping others escape the predator, while others play alone and attempt to escape by themselves [@Cere.etal2021]. Predators must learn how the prey behave and then decide how best to capture them. Thus, *DBD* simulates a highly dynamic system where both predators and prey must adjust to each other, suggesting that virtual systems could be useful in identifying general ecological patterns, which might in turn help to advance the current gaps in predator-prey research. Other studies on virtual predator-prey systems show that predation regimes can drive individual variation in risk perception [@Beauchamp2020], that familiarity between prey has a positive indirect effect on survival [@Cere.etal2021], and that prey face contrasting natural and social selection regimes (Santostefano et al. in prep). Virtual predator-prey systems generate large volumes of data on interacting players throughout their lifetime in the game under realistic ecological scenarios. Hence, they offer the opportunity to tackle fundamental questions about the role of experience and prey behaviour on individual predator foraging specialization along with their potential fitness consequences.

In this study, we test how hunting experience shapes predator foraging specialization using individual behavioural data from players in the online videogame *Dead by Daylight*. *DBD* simulates a direct predator-prey interaction, where one predator player hunts four prey players in different virtual environments. The data grants a high degree of precision on the behavioural interaction, as the behaviour of both the predator and the four prey along with the predator's success are monitored simultaneously in each trial. First, we investigate how predators develop their individual hunting expertise. We hypothesize that predators should differ in the development of their expertise, partly because they encounter varying levels of difficulty with the prey that they pursue. For example, a predator may face greater difficulty than other individuals if it more often encountered prey that were elusive. Second, we test the hypothesis that experience will shape foraging specialization. If experience reduces the costs of switching between hunting tactics, we predict that the predator population should become more flexible. Alternatively, if experience enables the refinement of the hunting tactics, then the predator population may instead specialize. Otherwise, if both mechanisms operate at the same time, then we should see no change with experience at the population level, but differences in specialization among individuals should increase. Third, we evaluate how predator foraging specialization emerges from behavioural interactions with prey. Whether predators specialize or not with experience should depend on the behaviour of their prey. We expect that predators that experienced more predictable encounters with their prey will specialize, while predators that experienced unpredictable encounters with their prey should adopt a flexible hunting strategy. If we detect such prey-dependent fine-tuning, then specialist and flexible hunters should attain equal success.

@@

*DBD* is a survival asymmetric (i.e. a game where the gameplay  mechanics differ between two groups) multiplayer online game developed by Behaviour Interactive Inc, in which players can play either as a predator or a prey. The objective of the predator is to hunt and capture the four prey across a virtual environment. The objective of the four prey is to search for resources while avoiding the predator. The resources are in the form of power generators that, once all activated, will enable the prey to escape through one of two exit doors. The composition of the predator and prey group for a match is determined by a skill-based matchmaking algorithm. A match ends when the predator kills all the prey available (i.e. that have not escaped), or when the last remaining prey escapes the virtual environment. Each player, predator or prey, can choose an avatar with abilities that encourage specific play styles (e.g. bold vs cautious prey, or ambush vs roaming predator). During our study period, the game offered 23 predator avatars. The virtual environment where matches take place is composed of fixed and procedurally generated habitat components, such as vegetation, mazes, and buildings. Some of these environments are larger than others, with varying structural complexity. However, predators display only minimal changes in behaviour and hunting success across the environments, probably due to a game feature enabling them to have visual cues of the generators [@FraserFranco.etal2022]. There were `r toString(format(length(unique(dat$environment_id))))` virtual game environments available for play during our study period.

@@

The videogame company provided data that spanned a period of 6 months of gameplay recorded for every player. The first recorded match was played on `r toString(format(min(dat$date)))` and the last one on `r toString(format(max(dat$date)))`. We cleaned and filtered the raw data to produce a dataset appropriate for our analyses. We analyzed only matches where players did not know each other (i.e. "Online" mode). We filtered any matches where players were inactive, such as when speed values were equal to, or very close to, zero. Moreover, we used our knowledge of the game to remove any matches where players were potentially hacking, or not playing the game as it is supposed to be played. We then sampled players that played 300 matches or more. To optimize the representation of gameplay experience at advanced levels, we set the maximum number of matches at 500 for these players. For instance, there could be a large difference in gameplay between 300-500 and 500-1000 matches. Players that played more than 500 matches represented ~2% of the population.

Our population consists of `r toString(format(length(unique(dat$predator_id))))` predator players with a total record of `r toString(formatC(length(unique(dat$match_id)), big.mark = " "))` matches. The predator-players' experience varied between `r toString(format(min(dat$total_xp_pred)))` and `r toString(format(max(dat$total_xp_pred)))` matches played. These matches lasted between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)`-min (mean = `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)`-min). The following information is collected and reported for every match : the player's anonymous ID, its avatar (i.e. the predator character chosen with its specific powers-gameplay mechanics), the game environment, the predator-player's experience along with its speed, and the average speed of the group of prey it encountered.

We recognize that we could have introduced a bias in our analyses since we sampled individuals who stayed in the game for at least 300 matches. For instance, these players may already be used to play videogames, and thus, could differ in the way they play at early stages of their experience in the game. To ensure that our sample was not biased, the videogame company provided a random sample of players that played between 20 and 50 matches, between 51 and 100 matches, and between 101 and 300 matches from the same timeframe as our current sample. We then took the first 20 matches played by these players, including those from our sampled population, and compared their average behaviour using a Bayesian hierarchical linear model. We found that neither of the four groups differed in their average speed (Table SX), which gives us confidence that our player population was not biased.

@@

We analyzed the predator's average speed and the average speed of the prey group encountered by the predator. The predator's average speed is measured as the average distance traveled per second during a match ($\overline{x}$ = `r toString(format(round(mean(dat$pred_speed), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$pred_speed), digits = 2), nsmall = 2))` m/s). We measured the preys' average speed as the average travel speed of the four individual prey within a match ($\overline{x}$ = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s). We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). Lastly, we defined the predator's cumulative experience as the number of matches played prior to the match being monitored. For example, the first match of a player would have a cumulative experience value of 0, while the tenth match would have a value of 9.

@@

We used the default number of basis functions (K) in "brms" for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id$) followed a Gaussian distribution with estimated standard deviation ($id \sim N(0,\ I\ \sigma_{id})$). We used Gaussian priors for the game duration ($N(0,\ 1)$), the intercept ($N(0,\ 0.5)$), and the precision paremeter ($N(2,\ 0.5)$). We employed the default Student t priors for the smoothing parameter ($t(3,\ 0,\ 2.5)$). After fitting the three models, we proceeded to select the one with the best predictive accuracy using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@Vehtari.etal2017;@Piironen.Vehtari2017;@Vehtari.etal2022].

### Changes in foraging behaviour and predator-prey interactions with experience

We tested whether foraging behaviour changed with experience, and compared the success of generalists vs specialists, using a multivariate, double-hierarchical generalized linear model (MDHGLM) [@leeDoubleHierarchicalGeneralized2006;@cleasbyQuantifyingPredictabilityBehaviour2015;@westneatBiologyHiddenResidual2014;@Mitchell.etal2016a;@ODea.etal2022]. We categorized the predators by their experience based on the number of matches they played (i.e. their cumulative experience). Since we monitored all predator players throughout their experience, they appeared in all of the three following experience categories. First, we assigned predators the status of "novice" when they had a cumulative experience below 100 matches. For example, a predator player that played 500 matches in total would be labelled as a novice in its first 100 matches. Following this logic, we labelled predators as "intermediate" hunters when they had a cumulative experience between 100 and 299 and matches. Lastly, we defined predators as "advanced" hunters when they had a cumulative experience equal or above 300, with a maximum of 499 cumulated matches.

For each level of experience (i.e. novice, intermediate, advanced), the model estimated the mean speed of every individual (which we call the mean part of the model) and, by using a heterogeneous structure of the residuals, the intra-individual standard deviation of speed for every individual (which we call the dispersion part of the model). Thus, in the mean part of the model, we could estimate among individual differences in the foraging tactic used by predators (i.e. along the slow-fast continuum), while in the dispersion part of the model, we could estimate among individual differences in foraging specialization/flexibility (i.e. intra-individual variability). We followed the same structure for the preys' speed to estimate among individual differences in the mean and standard deviation of speed of the prey groups encountered by predators. For hunting success, we only modeled the mean part of the equation to estimate among individual differences in mean hunting success. For the predator and the prey's speed, we controlled for the average rank of the prey that they encountered as proxy for prey skills on both the mean and dispersion part of the equation. The ranking system in *DBD* was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank). We used the mean rank of the four prey and applied a square-root transformation. For the predator's hunting success, we controlled for the square-root of the duration of the match.

We estimated the correlation between the individual random effect on the mean and on the dispersion within each trait (e.g. $cor(\sigma_{id_{m}\ pred}\ \sigma_{id_{d}\ pred})$) and among each trait (e.g. $cor(\sigma_{id_{m}\ pred}\ \sigma_{id_{d}\ prey})$) within a level of experience. In the first case, this allows us to assess, for example, whether predators that are on average slower/faster are more specialized or flexible. In the second case, the correlations enables us to estimate, for example, whether predators that encountered prey that were faster/slower adopted a slower/faster (or more specialized/flexible) tactic, or if slower/faster predators had lower/greater hunting success. As a complement, the model also estimated correlations among the different levels of experience (e.g. $cor(\sigma_{id_{m}\ pred\ novice}\ \sigma_{id_{d}\ pred\ intermediate})$). The latter "character-state" approach [@Via.Lande1985;@viaAdaptivePhenotypicPlasticity1995], enables us to investigate whether predators express similar behaviours across each of three developmental stages. We present the character-state analyses as additional material in the Appendix S2 as they were not part of our initial objectives.

@@

We parameterized equations (1) and (2) with a Gaussian distribution family (predator and prey speed) and used a beta-binomial distribution family for equation (3) (predator hunting success). We used Gaussian priors for the prey rank ($N(2.5,\ 1)$) and game duration ($N(10,\ 2)$) on the mean part of the model, and the default priors in "brms" for the dispersion part. We used the default Student t priors in "brms" for the intercepts on the mean and dispersion part of the equation. For hunting success, we applied a Gaussian prior on the precision parameter ($N(2,\ 1)$). For every predicted variable, we applied a half-Gaussian on the random effects for the mean part of the model ($N(0,\ 1)$), and the default "brms" Student t prior for the dispersion part of the model.

### Markov Chain Monte Carlo settings

We parametrized every model described above to run four MCMC chains with 1000 posterior samples for each parameter. We ran 2500 iterations with a thinning set to eight for the additive model with a global smoother only (see Table I), and 1500 iterations with a thinning set to four for the additive model with a global smoother and group-level smoothers as well as for the additive model with group-level smoothers only (Table I). Burn-in was set to 500 iterations in each model. We parametrized the MDHGLM to run 2500 iterations with a thinning set to 8, with the first 500 iterations used as warmups. We assessed the convergence of the MCMC chains using trace plots, R-hat diagnostics with a threshold of <1.05, and effective sample sizes (ESS) with a threshold of >100 for the bulk-ESS and tail-ESS [@Vehtari.etal2021a]. We also performed posterior predictive checks which showed an adequate fit of the models (for details, see https://github.com/quantitative-ecologist/experience-hunting-tactics).

@@

# ACKNOWLEDGEMENTS
